Galaxy Imaging with Generative Models: Insights from a Two-Models Framework

### 2025-03-16 MNRAS/galaxy-generative-models-overleaf-submit-2nd/article_MNRAS_r1.tex
before 1st revision submission

\section{Introduction}
\label{sec:Intro}
Image generation in Machine Learning (ML) is a challenging task, particularly in the context of non-ergodic processes. Recently, significant breakthroughs in image quality have been achieved thanks to large-scale statistical model architectures such as \texttt{DALL-E2} \citep{ramesh2022}, \texttt{Midjourney} \citep{Oppenlaender2022}, and \texttt{StableDiffusion} \citep{Rombach2022}, which leverage \textit{stochastic diffusion processes}. These models have largely replaced the previous generation of approaches based on \textit{variational autoencoders} (VAE) \citep{Kingma2014}\footnote{Note: Updated in 2022.}, \textit{adversarial networks} (GAN) \citep{goodfellow2014generative} as employed in works such as \citep[e.g.,][]{KarrasALL18,brock2018large}, and \textit{normalizing flows}, exemplified by \texttt{Glow} \citep{Kingma2018}.

@@

{\color{red}The central challenge in generation lies in estimating the \textit{score}, the key ingredient of the generation process.} Two main approaches have been proposed: (1) deriving a model for $p_t$ that incorporates regularity patterns and correlations to learn the \textit{score} via \textit{score matching} \citep{hyvarinen2005a}, or (2) employing a \textit{denoising} neural network, as introduced by \cite{Bengio2013}, further developed in \cite{Sohl-Dickstein2015,Ho2020,song2021scorebased,song2021maximum}, and recently examined from a mathematical perspective in \cite{kadkhodaie2024generalization}. The first approach has been utilized in works such as \cite{Guth2022b,Lempereur2024} with Gibbs energy parameterization. The second has found applications in astrophysics, including galaxy image generation \citep{smith2021}, generation of 21~cm luminosity temperature maps \citep{Zhao2023}, super-resolution of large-scale cosmic structures \citep{Schanz2023}, and Bayesian posterior sampling for weak lensing mass-mapping problems \citep{Remy2023}.

@@

Returning to the original GAN problem, the question \textit{"Does $p_G = p_{\text{data}}$?"} remains an open challenge. This is why numerical experiments are still a valuable approach. We have selected three models, each representing one of the three types described in this section, to assess the reliability of model optimization. This choice may be questioned but we think that it serves more as guide lines if one would like to use other models. 
Regarding the number of model parameters, we maintain them in the range of $O(10)$ million, which is significantly lower than the scale used in large diffusion models mentioned in the introduction. This choice is relevant for two reasons: first, it aligns with the constraints faced by practitioners with limited computational resources; second, it is well-suited to the sizes of the images ($d \times d$) and the dataset ($N$) used in our experiments.  
Anticipating one of the results, we observe that, as a general rule of thumb, the ratio of the total number of available training pixels to the total number of model parameters should be at least on the order of $O(10^2)$ to ensure effective learning.
}

@@

A partial answer to these questions is provided in \cite{kadkhodaie2024generalization} within the context of diffusion-based models, highlighting the impact of dataset size on the transition from \textit{memorization} to \textit{generation} regimes. Following their methodology, we compare two generative models of the same architecture trained on non-overlapping subsets of the same dataset. For simplicity, we limit our experiment to three models and a single dataset, which we believe suffices to derive meaningful insights. 
{\color{red}We focus on investigating how one can perform tests, particularly with the two-model scenario, to ensure that the model behaves as intended. 
}

@@

{\color{red}The author have perform some tests with other settings of the above architectures and even if not exhaustive, the results presented in the following section are generic and would serve as guide line if one would like to use either another settings of the models I have used, either to test a completely different model architecture.

@@

{\color{red}
In the present study, we focus on galaxy image generation, taking as a reference the experiment conducted by \cite{kadkhodaie2024generalization}, who, however, used widely known machine learning datasets as face or bedroom images. All their images were grayscale (single channel), with intensity values in the range $[0,1]$ and a resolution mostly of $80 \times 80$ pixels.

We use the Sloan Digital Sky Survey (SDSS) Data Release 9 dataset, as described in \citep{smith2021}\footnote{The script to download the original selected dataset is available at \url{https://github.com/Smith42/astroddpm/data/sdss}.}. As the authors mentioned, the volume complete sample includes $(g,r,z)$ bands, with an $r$-band absolute magnitude limit of $M_r\leq 20$ and a redshift limit of $z\leq 0.08$. Each band image consists of $512\times 512$ pixels at a resolution of $0.262^{\prime\prime} \mathrm{pixel}^{-1}$ (aka \texttt{pixscale}), with an average seeing of approximately $1^{\prime\prime}$. Our preprocessing applied to the downloaded dataset, consisting of approximately $250,000$ images, includes cropping around the target galaxy coordinates to $64 \times 64$ pixels to ensure that each image predominantly contains a single object. Additionally, a single grayscale channel was created from the $(z, g, r)$ bands using the \texttt{scipy} \citep{2020SciPy-NMeth} library to balance the contributions of the different bands\footnote{The \texttt{make\_lupton\_rgb} function was applied with settings $Q=8$ and \texttt{stretch}=0.2.}. Of course, this choice of dataset is by no means restrictive, and researchers may use their own datasets. To facilitate reproducibility, we provide additional materials in the companion GitHub repository, enabling users to replicate the experiments.

Thus, we restrict ourselves to relatively modest image sizes compared to those produced by models such as \texttt{light-weight-gan} and \texttt{Glow}. Our primary objective is to investigate whether the transition from \textit{memorization} to \textit{generalization} in diffusion models occurs around $10^5$ training images, as mentioned earlier, and whether this transition aligns with the number of model parameters (see Section~\ref{sec-some-comments}). Furthermore, we aim to examine how GAN and flow-based models behave as a function of dataset size.

@@

The distributions of morphological coefficients derived from the three models show good agreement with those computed from the original images. However, we do not aim for high-quality images or state-of-the-art results, as seen in \citep{ravanbakhsh2016, Fussell2019, Lanusse2021, smith2021, HACKSTEIN2023100685}, which achieve better image quality. Instead, our focus is on a different question: with low-resolution and relatively simple images, do models of the same type trained on moderately sized datasets learn the same underlying probability distribution? Furthermore, if differences exist, can we identify measures to assess the reliability of the results?
%
\subsubsection{\texttt{U-Net} Denoising Performances}
\label{sec-UNet-Perf}
%
%Before assessing the generative performances of the models, we first present results regarding the denoising performances of two \texttt{U-Net} networks.
{\color{red} Before assessing the generative performance of the diffusion models in the next section, we first present the denoising performance of two \texttt{U-Net} networks with the same architecture, which are subsequently used to estimate the score as discussed in Section~\ref{sec-diff} (Equation~\ref{eq-backward-diffusion}).
}

@@

In conclusion, while modern generative models can produce high-quality galaxy images with reasonably sized datasets, much remains to be explored regarding the mechanisms governing their learning processes and the factors influencing their generalization.  
{\color{red}
As a key takeaway, we first highlight the necessity of sufficiently large training datasets to ensure effective learning. The exact amount of data required is not yet settled and remains an open research question, but it certainly depends on the image resolution, the number of model parameters, and the model architecture. Investigating different strategies to improve generalization without requiring massive datasets remains an open question. In future work, one could explore restricting the diversity of galaxies in the dataset or leveraging pre-trained models on general-purpose images to enhance training efficiency.  
Secondly, we emphasize the importance of evaluations beyond morphological distributions to properly assess whether model optimization is effective. The "two-models" methodology proved to be a valuable approach, enabling us to assess both the similarity of generated images to real-world data and the consistency of the learned probability distributions. This framework has been a key element of our study.  
Finally, we argue that score-based diffusion models offer a direct way to test whether they operate in the generalization regime.

>>>>

\section{Introduction}
\label{sec:Intro}
Image generation in Machine Learning (ML) is a challenging task, particularly in the context of non-ergodic processes. Recently, significant breakthroughs in image quality have been achieved thanks to large-scale statistical model architectures such as \texttt{DALL-E2} \citep{ramesh2022}, \texttt{Midjourney} \citep{Oppenlaender2022}, and \texttt{StableDiffusion} \citep{Rombach2022}, which leverage \textit{stochastic diffusion processes}. These models have largely replaced the previous generation of approaches based on \textit{variational autoencoders} (VAE) \citep{Kingma2014}\footnote{Note: Updated in 2022.}, \textit{adversarial networks} (GAN) \citep{goodfellow2014generative} as employed in works such as \citep[e.g.,][]{KarrasALL18,Brock2019}, and \textit{normalizing flows}, exemplified by \texttt{Glow} \citep{Kingma2018}.

@@

{\color{red}The central challenge in generation lies in estimating the \textit{score}, the key ingredient of the generation process.} Two main approaches have been proposed: (1) deriving a model for $p_t$ that incorporates regularity patterns and correlations to learn the \textit{score} via \textit{score matching} \citep{hyvarinen2005a}, or (2) employing a \textit{denoising} neural network, as introduced by \cite{Bengio2013}, further developed in \cite{Sohl-Dickstein2015,Ho2020,song2021scorebased,song2021maximum}, and recently examined from a mathematical perspective in \cite{kadkhodaie2024generalization}. The first approach has been utilized in works such as \cite{guth2022wavelet,Lempereur2024} with Gibbs energy parameterization. The second has found applications in astrophysics, including galaxy image generation \citep{smith2021}, generation of 21~cm luminosity temperature maps \citep{Zhao2023}, super-resolution of large-scale cosmic structures \citep{Schanz2023}, and Bayesian posterior sampling for weak lensing mass-mapping problems \citep{Remy2023}.

@@

Returning to the original GAN problem, the question \textit{"Does $p_G = p_{\text{data}}$?"} remains an open challenge. This is why numerical experiments are still a valuable approach. We have selected three models, each representing one of the three types described in the previous sections, to assess the reliability of model optimization. This choice may be questioned but we think that it serves more as guide lines if one would like to use other models. 
Regarding the number of model parameters, we maintain them in the range of $O(10)$ million, which is significantly lower than the scale used in large diffusion models mentioned in the introduction. This choice is relevant for two reasons: first, it aligns with the constraints faced by practitioners with limited computational resources; second, it is well-suited to the sizes of the images ($d \times d$) and the dataset ($N$) used in our experiments.  
Anticipating one of the results, we observe that, as a general rule of thumb, the ratio of the total number of available training pixels to the total number of model parameters should be at least on the order of $O(10^2)$ to ensure effective learning.
}

@@

A partial answer to these questions is provided in \cite{kadkhodaie2024generalization} within the context of diffusion-based models, highlighting the impact of dataset size on the transition from \textit{memorization} to \textit{generation} regimes. Following their methodology, we compare two generative models of the same architecture trained on non-overlapping subsets of the same dataset. For simplicity, we limit our experiment to three models and a single dataset, which we believe suffices to derive meaningful insights. 
{\color{red}We focus on investigating how one can perform tests to ensure that the model behaves as intended. 
}

@@

{\color{red}We have perform some tests with other settings of the above architectures and even if not exhaustive, the results presented in the following section are generic and would serve as guide line if one would like to use either another settings of the models we have used, either to test a completely different model architecture.

@@

{\color{red}
In the present study, we focus on galaxy image generation, taking as a reference the experiment conducted by \cite{kadkhodaie2024generalization}, who, however, used widely known machine learning datasets as face or bedroom images. For completeness, all their images were grayscale (single channel), with intensity values in the range $[0,1]$ and a resolution mostly of $80 \times 80$ pixels.

We use the Sloan Digital Sky Survey (SDSS) Data Release 9 dataset, as described in \citep{smith2021}\footnote{The script to download the original selected dataset is available at \url{https://github.com/Smith42/astroddpm/data/sdss}.}. As the authors mentioned, the volume complete sample includes $(g,r,z)$ bands, with an $r$-band absolute magnitude limit of $M_r\leq 20$ and a redshift limit of $z\leq 0.08$. Each band image consists of $512\times 512$ pixels at a resolution of $0.262^{\prime\prime} \mathrm{pixel}^{-1}$, with an average seeing of approximately $1^{\prime\prime}$. Our preprocessing applied to the downloaded dataset, consisting of approximately $300,000$ images, includes cropping around the target galaxy coordinates to $64 \times 64$ pixels to ensure that each image predominantly contains a single object. Additionally, a single grayscale channel was created from the $(z, g, r)$ bands using the \texttt{scipy} \citep{2020SciPy-NMeth} library to balance the contributions of the different bands\footnote{The \texttt{make\_lupton\_rgb} function was applied with settings $Q=8$ and \texttt{stretch}=0.2.}. Of course, this choice of dataset is by no means restrictive, and researchers may use their own datasets. To facilitate reproducibility, we provide additional materials in the companion GitHub repository, enabling users to replicate the experiments.

Thus, we restrict ourselves to relatively modest image sizes compared to those produced by models such as \texttt{light-weight-gan} and \texttt{Glow}. Our primary objective is to investigate whether the transition from \textit{memorization} to \textit{generalization} in diffusion models occurs around $10^5$ training images, as mentioned earlier (see Section~\ref{sec-some-comments}). Furthermore, we aim to examine if similar transition occurs for  GANs and flow-based models.

@@

The distributions of morphological coefficients derived from the three models show good agreement with those computed from the original images. However, we do not aim for high-quality images or state-of-the-art results, as seen in \citep{ravanbakhsh2016, Fussell2019, Lanusse2021, smith2021, HACKSTEIN2023100685}, which achieve better image quality. Instead, our focus is on a different question: with low-resolution and relatively simple images, do models of the same type trained on moderately sized datasets learn the same underlying probability distribution? Furthermore, if differences exist, can we identify measures to assess the reliability of the results?
%
\subsubsection{\texttt{U-Net} Denoising Performances}
\label{sec-UNet-Perf}
%
%Before assessing the generative performances of the models, we first present results regarding the denoising performances of two \texttt{U-Net} networks.
{\color{red} Before assessing the generative performance of the diffusion models in the next section, we first present the denoising performance of two \texttt{U-Net} networks with the same architecture, which are subsequently used to estimate the score as discussed in Section~\ref{sec-diff} (Equations~\ref{eq-backward-diffusion},\ref{thm-score-denoising}).
}

@@

In conclusion, while modern generative models can produce high-quality galaxy images with reasonably sized datasets, much remains to be explored regarding the mechanisms governing their learning processes and the factors influencing their generalization.  
{\color{red}
As a key takeaway, we first highlight the necessity of sufficiently large training datasets to ensure effective learning. The exact amount of data required is not yet settled and remains an open research question, but it certainly depends on the image resolution, the number of model parameters, and the model architecture. Investigating different strategies to improve generalization without requiring massive datasets remains an open question. In future work, one could explore for instance restricting the diversity of galaxies in the dataset or leveraging pre-trained models on general-purpose images to enhance training efficiency.  
Secondly, we emphasize the importance of evaluations beyond morphological distributions to properly assess whether model optimization is effective. The "two-models" methodology proved to be a valuable approach, enabling us to assess both the similarity of generated images to real-world data and the consistency of the learned probability distributions. This framework has been a key element of our study.  
Finally, we argue that score-based diffusion models offer a direct way to test whether they operate in the generalization regime.
